---
title: "Time Series Forecasting"
output: html_document
date: "2025-01-02"
---

#Section 1: Data Import and Initial Exploration

# 1.1 Data Import
```{r}
tsf <- read.csv("C:/Users/jivko/Documents/Data Analytics, Big Data, and Predictive Analytics/Personal Project/Time Series Forecasting - Gold/gold_price_data.csv")

# Print the first few rows
head(tsf, 40)
```

# 1.2 Convert the Timestamp Column to Date Format
```{r}
# Convert the Timestamp column to Date format
tsf$Date <- as.Date(tsf$Date)
```


# 1.3 Data Structure Overview
```{r}
str(tsf)
```

# 1.4 Plotting the Time Series
```{r}
# Plot the gold price over time
plot(tsf$Date, tsf$Value, type="l", xlab="Date", ylab="Gold Price", main="Gold Price Time Series")
```

#Section 2: Handling Date Differences and Uniform Resampling

# 2.1 Calculating Differences Between Consecutive Dates
```{r}
# Calculate differences between consecutive dates
date_diff <- diff(tsf$Date)

# Summary of differences
summary(date_diff)
```
# 2.2 Unique Time Differences Count
```{r}
# View unique time differences and their counts
table(date_diff)
```
# 2.3 Visualizing Time Differences
```{r}
# Plot histogram of time differences
hist(as.numeric(date_diff), main="Distribution of Time Differences", xlab="Time Differences")
```
# 2.4 Resampling Data to a Uniform Frequency
```{r}
# Load required library
library(zoo)

# Create a complete sequence of dates
full_dates <- seq(min(tsf$Date), max(tsf$Date), by="day")

# Merge with your dataset, filling missing dates with NA
tsf_complete <- merge(data.frame(Date=full_dates), tsf, by="Date", all.x=TRUE)

# Interpolate missing values
tsf_complete$Value <- na.approx(tsf_complete$Value, na.rm=FALSE)
```

```{r}
# Optionally, if you want to save the table as a CSV file
write.csv(tsf_complete, "tsf_complete.csv", row.names = FALSE)
```


# 2.5 Verification of Complete Daily Sequence
```{r}
# Recalculate differences and check if all are 1 day
all(diff(tsf_complete$Date) == 1)  # Should return TRUE
```
```{r}
# Calculate differences between consecutive dates
date_diff2 <- diff(tsf_complete$Date)

# Summary of differences
summary(date_diff2)
```
```{r}
# View unique time differences and their counts
table(date_diff2)
```
#Section 3: Interpretation of ADF Test Results

# 3.1 ADF Test for the Original Series
```{r}
# Install and load the tseries package (if not already installed)
if (!require(tseries)) install.packages("tseries")
library(tseries)

# Perform the ADF Test
adf_test <- adf.test(tsf_complete$Value, alternative = "stationary")

# Display results
print(adf_test)
```
# 3.2 ADF Test for the Differenced Series
```{r}
# Apply first-order differencing
tsf_diff <- diff(tsf_complete$Value)

# Perform ADF test again
adf_diff_test <- adf.test(tsf_diff, alternative = "stationary")
print(adf_diff_test)
```

#Section 4: ARIMA Model Fitting and Forecasting

# 4.1 Fitting an ARIMA Model & Forecasting Future Values
```{r}
# Load necessary libraries
library(forecast)
library(tibble)

# Convert the data to a time series object (daily frequency)
ts_data <- ts(tsf_complete$Value, frequency = 365)

# Fit an ARIMA model
model <- auto.arima(ts_data)
summary(model)  # Display model summary

# Forecast the next 730 days (2 years)
forecasted_values <- forecast(model, h = 730)

# Generate a date sequence for the forecast period (next 2 years)
forecast_dates <- seq(from = max(tsf_complete$Date) + 1, by = "day", length.out = 730)

# Create a combined data frame with forecasted data
forecast_data <- data.frame(
  Date = forecast_dates,
  Forecast = round(forecasted_values$mean, 2),
  Lower_80 = round(forecasted_values$lower[,2], 2),
  Upper_80 = round(forecasted_values$upper[,2], 2),
  Lower_95 = round(forecasted_values$lower[,1], 2),
  Upper_95 = round(forecasted_values$upper[,1], 2)
)

# View forecast table
print(forecast_data)

# Optionally, if you want to save the table as a CSV file
write.csv(forecast_data, "forecast_table.csv", row.names = FALSE)
```

# 4.2 Visualization of Forecast Results
```{r}
# Load necessary libraries
library(forecast)
library(ggplot2)
library(dplyr)

# Convert the data to a time series object (daily frequency)
ts_data <- ts(tsf_complete$Value, frequency = 365)

# Fit an ARIMA model
model <- auto.arima(ts_data)
summary(model)  # Display model summary

# Forecast the next 730 days (2 years)
forecasted_values <- forecast(model, h = 730)

# Create a data frame for the original data
tsf_complete <- tsf_complete %>%
  mutate(Date = as.Date(Date))

# Generate a sequence of forecast dates for the next 2 years
forecast_dates <- seq(from = max(tsf_complete$Date) + 1, by = "day", length.out = 730)

# Prepare a data frame for forecasted data
forecast_df <- data.frame(
  Date = forecast_dates,
  Mean = forecasted_values$mean,
  Lower80 = forecasted_values$lower[, 1],
  Upper80 = forecasted_values$upper[, 1],
  Lower95 = forecasted_values$lower[, 2],
  Upper95 = forecasted_values$upper[, 2]
)

# Create ggplot
p <- ggplot() +
  geom_line(data = tsf_complete, aes(x = Date, y = Value, color = "Original Data"), size = 1, linetype = "solid") +  # Original data
  geom_line(data = forecast_df, aes(x = Date, y = Mean, color = "Forecast Mean"), size = 1.2) +  # Forecast mean
  geom_ribbon(data = forecast_df, aes(x = Date, ymin = Lower80, ymax = Upper80, fill = "80% CI"), alpha = 0.2) +  # 80% CI
  geom_ribbon(data = forecast_df, aes(x = Date, ymin = Lower95, ymax = Upper95, fill = "95% CI"), alpha = 0.1) +  # 95% CI
  labs(title = "Gold Price Time Series and Forecast - ARIMA Model",
       x = "Date",
       y = "Gold Price") +
  scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(name = "Legend", values = c("Original Data" = "black", "Forecast Mean" = "blue")) +
  scale_fill_manual(name = "Confidence Intervals", values = c("80% CI" = "blue", "95% CI" = "green")) +
  guides(color = guide_legend(title = "Legend"), fill = guide_legend(title = "Confidence Intervals"))

# Print the plot
print(p)

# Save the plot with increased width
ggsave("gold_price_forecast.pdf", plot = p, width = 12, height = 6)  # Save as PDF
```

```{r}
getwd()
```

#Section 5: SARIMA Model Forecasting

# 5.1 Fitting a SARIMA Model, & Forecasting Future Values, & Visualization of Forecast Results
```{r}
# Load necessary libraries
library(forecast)
library(ggplot2)
library(dplyr)

# Convert the data to a time series object (daily frequency)
ts_data <- ts(tsf_complete$Value, frequency = 365)

# Fit a SARIMA model (with seasonal component)
sarima_model <- auto.arima(ts_data, 
                           seasonal = TRUE,  # Set seasonal to TRUE
                           stepwise = FALSE, 
                           approximation = FALSE)

# Display the summary of the SARIMA model
summary(sarima_model)

# Forecast the next 730 days (2 years)
forecasted_values_sarima <- forecast(sarima_model, h = 730)

# Generate a sequence of forecast dates for the next 2 years
forecast_dates <- seq(from = max(tsf_complete$Date) + 1, by = "day", length.out = 730)

# Prepare a data frame for the forecasted data
forecast_df_sarima <- data.frame(
  Date = forecast_dates,
  Mean = forecasted_values_sarima$mean,
  Lower80 = forecasted_values_sarima$lower[, 1],
  Upper80 = forecasted_values_sarima$upper[, 1],
  Lower95 = forecasted_values_sarima$lower[, 2],
  Upper95 = forecasted_values_sarima$upper[, 2]
)

# Create ggplot for the SARIMA forecast
p_sarima <- ggplot() +
  geom_line(data = tsf_complete, aes(x = Date, y = Value, color = "Original Data"), size = 1, linetype = "solid") +  # Original data
  geom_line(data = forecast_df_sarima, aes(x = Date, y = Mean, color = "Forecast Mean"), size = 1.2) +  # Forecast mean
  geom_ribbon(data = forecast_df_sarima, aes(x = Date, ymin = Lower80, ymax = Upper80, fill = "80% CI"), alpha = 0.2) +  # 80% CI
  geom_ribbon(data = forecast_df_sarima, aes(x = Date, ymin = Lower95, ymax = Upper95, fill = "95% CI"), alpha = 0.1) +  # 95% CI
  labs(title = "Gold Price Time Series and Forecast - SARIMA Model",
       x = "Date",
       y = "Gold Price") +
  scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(name = "Legend", values = c("Original Data" = "black", "Forecast Mean" = "blue")) +
  scale_fill_manual(name = "Confidence Intervals", values = c("80% CI" = "blue", "95% CI" = "green")) +
  guides(color = guide_legend(title = "Legend"), fill = guide_legend(title = "Confidence Intervals"))

# Print the SARIMA forecast plot
print(p_sarima)

# Optionally, save the plot as a PDF
ggsave("sarima_forecast.pdf", plot = p_sarima, width = 12, height = 6)  # Save as PDF

# Optionally, save the forecast data to CSV
write.csv(forecast_df_sarima, "sarima_forecast_table.csv", row.names = FALSE)
```

# 5.2 Preparing the Forecast Data Frame
```{r}
# Display the forecast table in the console
print(forecast_df_sarima)
```












